Staged typed type-preserving tagless ``interpreters'', \emph{finally}

### 2007-03-26 tagless/tagless-final.tex
move tagless interp work to its own directory

\documentclass[preprint]{sigplanconf}
\usepackage{amsmath}
\usepackage{natbib1}
\usepackage{comment}

@@

@@

What are our contributions:
\begin{enumerate}
\item we interpret/compile/partially-evaluate a typed source language
   from within a typed language.  The requirements of the host type
   system are \jacques{not so onerous?  simple?},
\item a simple, clean implementation,
\item a nicely comparable Haskell/MetaOCaml version,
\item a functorial Symantics that deals with uniformly across all
evaluators, from interpretation to partial evaluation,
\item clear and easy extensibility to more features,
\item only closed terms can be evaluated.
\end{enumerate}

@@

\subsection{How to make encoding flexible: abstract the interpreter}
We have implemented the above in (Meta)OCaml and Haskell.  In Haskell,
the term `constructor' functions are defined as methods in a type class
Symantics (the name means that the class interface gives the syntax for
the source language and its instances give the semantics).

@@

Our sample code test1 now reads |app (lam (\x -> x)) (bool True)|,
whose inferred type is |Symantics repr => repr Bool|.
For example, we can encode the classical |power| function, as well
as the partial application |\x -> power x 7| as

@@

We can do the above in (Meta)OCaml as well, replacing type classes with
functors, via a module type.

@@

\begin{code}
  module EXR = EX(R);;
\end{code}

and so |EXR.test1 ()| evaluates to the untagged boolean value |true|.
In Haskell, we define a\\
|newtype R a = R a| and |instance Symantics R|
with the straightforward implementation. We must stress that |R| is
a |newtype|. Although it may look like a tag, it has no run-time
representation (erased during compilation) and pattern-matching
against |R| cannot ever fail and so is operationally an identity and
compiled that way.

@@

\section{Tagless compiler (aka, staged interpreter)}\label{compiler}
Staged code in OCaml... Don't show the Haskell code (too much)

\section{Tagless partial evaluator}\label{PE}

Surprisingly, it is also possible to write a partial evaluator!  The
partial evaluator does have tags: whether a value is static or dynamic.
The pattern-matching on this \emph{phase} tag is always exhaustive; we still
have no universal type and no tags for object types.  The partial
evaluator uses the previously mentioned compiler to `interpret' dynamic
terms.  In this sense, the partial evaluator is a modular extension to
the compiler.

Compare the output of some of the ctest with the corresponding
ptest. Use power?

One should note that the partial evaluator has fully reused the code
of the compiler. It is truly the `composition' of the interpreter and
the compiler.

@@

\begin{code}
    ('c, int     ) repr_pe = VI of int
                           | E of ('c, int) code
    ('c, bool    ) repr_pe = VB of bool
                           | E of ('c, bool) code
    ('c, 'a -> 'b) repr_pe = VF of ('c, 'a) repr -> ('c, 'b) repr
                           | E of ('c, 'a -> 'b) code
\end{code}

@@

So, the full signature of Symantics looks like (cf Section 2)

\begin{code}
module type Symantics = sig
  type ('c,'sv,'dv) repr
  val int  : int  -> ('c,int,int) repr
  val bool : bool -> ('c,bool,bool) repr
  val add  : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
  val mul  : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
  val leq  : ('c,int,int) repr -> ('c,int,int) repr -> ('c,bool,bool) repr
  val if_ : ('c,bool,bool) repr ->
             (unit -> ('c,'sa,'da) repr) ->
             (unit -> ('c,'sa,'da) repr) -> ('c,'sa,'da) repr

  val lam : (('c,'sa,'da) repr -> ('c,'sb,'db) repr)
    -> ('c,(('c,'sa,'da) repr -> ('c,'sb,'db) repr),'da->'db) repr
  val app : ('c,(('c,'sa,'da) repr -> ('c,'sb,'db) repr),'da->'db) repr
    -> ('c,'sa,'da) repr -> ('c,'sb,'db) repr
  val fix : (('c,(('c,'sa,'da) repr -> ('c,'sb,'db) repr) as 's,'da->'db) repr 
             -> ('c,'s,'da->'db) repr)  -> ('c,'s,'da->'db) repr
end
\end{code}

@@

\begin{code}
  (Int -> r Int) ->
  (Bool -> r Bool) ->
  (forall alpha beta. (r alpha -> r beta) -> r (alpha->beta)) -> ... r tau
\end{code}

@@

Taha et al define a self-interpreter si to be an object-language term
such that, for any object-language term e,

@@

Walid's PEPM07 (Concoqtion paper)? At leats at the Concoqtion
presentation at WG2.11, Emir used tagless interpreter as the main and
only example.

@@

Regarding our implementation of the type-level function |'sv -> 'dv|.
Gibbons et al paper on
the implementations of typecase may be relevant. I have not read it,
but I have the feeling some of their typecase designs may be
relevant (at least to cite). 
However, the typecase paper relies on having a 
bijection between types -- our |'sv| and |'dv| in this case, which is not 
our situation.  We have a one-sided view of things, where |'sv -> 'dv| is 
easy, and |'dv| "perfectly reflects" an |'sv|, but you still can't get at 
that |'sv|.

@@

>>>>

\documentclass[preprint]{sigplanconf}
\usepackage{amsmath,amssymb}
\usepackage{natbib1}
\usepackage{comment}

@@

\newcommand{\ZZ}{\mathbb{Z}}
\newcommand{\BB}{\mathbb{B}}

@@

What are our contributions:
\begin{enumerate}
\item we interpret/compile/partially-evaluate a typed source language
   from within a typed language.  The requirements of the host type
   system are \jacques{not so onerous?  simple?},
\item a simple, clean implementation,
\item a nicely comparable Haskell/MetaOCaml version,
\item a functorial Symantics that deals with uniformly across all
evaluators, from interpretation to partial evaluation,
\item clear and easy extensibility to more features,
\item only closed terms can be evaluated.
\item an approach to self-interpretation which is compatible with the
  above.
\end{enumerate}

@@

\subsection{How to make encoding flexible: abstract the interpreter}
\label{encoding}
We have implemented the above in (Meta)OCaml and Haskell.  In Haskell,
the term `constructor' functions are defined as methods in a type class
Symantics (the name means that the class interface gives the syntax for
the source language and its instances give the semantics).

@@

|test1| now reads |app (lam (\x -> x)) (bool True)|,
whose inferred type is |Symantics repr => repr Bool|.
For example, we can encode the classical |power| function, as well
as the partial application |\x -> power x 7| as

@@

We can do the above in (Meta)OCaml as well, replacing type classes with
functors, via a module type.  For this section, the following simple
signature will be sufficient:

@@

\noindent
\begin{code}
  module EXR = EX(R);;
\end{code}

\noindent and so |EXR.test1 ()| evaluates to the untagged boolean value |true|.
In Haskell, we define a\\
|newtype R a = R a| and |instance Symantics R|
with the straightforward implementation. We must stress that |R| is
a |newtype|. Although it may look like a tag, it has no run-time
representation (erased during compilation) and pattern-matching
against |R| cannot ever fail and so is operationally an identity and
compiled that way.

@@

\section{Tagless compiler (aka, staged interpreter)}\label{compiler}

We can also write a ``compiler'' from our lambda calculus directly
into OCaml code, using MetaOCaml's staging facilities.  This compiler
attempts no optimizations at all, it simply produces residual code.
This is a completely straightforward staging of |module R|.

% taken from incope.ml -- should this be put into another file, to make
% sure the paper does not get out of synch?
\begin{code}
module C = struct
  type ('c,'sv,'dv) repr = ('c,'dv) code
  let int (x:int) = .<x>.
  let bool (b:bool) = .<b>.
  let add e1 e2 = .<.~e1 + .~e2>.
  let mul e1 e2 = .<.~e1 * .~e2>.
  let leq x y = .< .~x <= .~y >.
  let eql x y = .< .~x = .~y >.
  let if_ eb et ee = 
    .<if .~eb then .~(et () ) else .~(ee () )>.

  let lam f = .<fun x -> .~(f .<x>.)>.
  let app e1 e2 = .<.~e1 .~e2>.
  let fix f = 
     .<let rec self n = .~(f .<self>.) n in self>.
  let dyn_id (x:('c,'sv,'dv) repr):('c,'sv1,'dv) repr = x
end;;
\end{code}

\noindent It is readily obvious that this compiler does not suffer
any interpretive overhead whatsoever, nor does the resulting code
contain any tags.  We add one item to the language, |dyn_id|, as it
is needed in the next section, where we will explain its purpose.

\jacques{we need to say something about the Haskell version here too,
even though we don't show it.  This is the first place we use GADT,
so maybe the ByteCode data type should be shown?}
\section{Tagless partial evaluator}\label{PE}

Surprisingly, it is also possible to write a partial evaluator!  The
partial evaluator does have tags: whether a value is static or dynamic.
The pattern-matching on this \emph{phase} tag is always exhaustive; we still
have no universal type and no tags for object types.  The partial
evaluator uses the previously mentioned compiler to ``interpret'' dynamic
terms, and it uses the interpreter to reduce static terms.  In this sense, 
the partial evaluator is a modular extension to both the compiler and
the interpreter.

\subsection{MetaOCaml version}

Here is the full code for the MetaOCaml version.
\begin{code}
module P = struct
  type ('c,'sv,'dv) repr = {st: 'sv option; dy: ('c,'dv) code}
  let abstr {dy = x} = x
  let pdyn x = {st = None; dy = x}

  let int  (x:int)  = {st = Some (R.int x); dy = C.int x}
  let bool (x:bool) = {st = Some (R.bool x); dy = C.bool x}

  (* generic build - takes a repr constructor, an 
     interpreter function and a compiler function (all 
     binary) and builds a PE version *)
  let build cast f1 f2 = fun e1 -> fun e2 -> 
     match (e1,e2) with
         ({st = Some n1}, {st = Some n2}) -> cast (f1 n1 n2)
        | _ -> pdyn (f2 (abstr e1) (abstr e2))
  (* same as 'build' but takes care of the neutral element 
     (e) simplification allowed via a monoid structure which is 
     implicitly present *)
  let buildsimp cast e f1 f2 = fun e1 e2 -> 
     match (e1,e2) with
        | ({st = Some e'}, _) when e = e' -> e2
        | (_, {st = Some e'}) when e = e' -> e1
        | ({st = Some n1}, {st = Some n2}) -> cast (f1 n1 n2)
        | _ -> pdyn (f2 (abstr e1) (abstr e2))

  let add e1 e2 = buildsimp int 0 R.add C.add e1 e2
  let mul e1 e2 = buildsimp int 1 R.mul C.mul e1 e2
  let leq e1 e2 = build bool R.leq C.leq e1 e2
  let eql e1 e2 = build bool R.eql C.eql e1 e2
  let if_ eb et ee = match eb with
           {st = Some b} -> if b then et () else ee ()
         | _ -> pdyn (C.if_ (abstr eb) 
                          (fun () -> abstr (et ()))
                          (fun () -> abstr (ee ())))
  let lam f = {st = Some f; 
               dy = C.lam (fun x -> abstr (f (pdyn x)))}
  let app ef ea = match ef with
        {st = Some f} -> f ea
      | _ -> pdyn (C.app (C.dyn_id (abstr ef)) (abstr ea))

   (* For now, to avoid divergence at the PE stage,
      we unroll the fixpoint exactly once, and then
      residualize *)
  let fix f = f (pdyn (C.fix (fun x -> abstr (f (pdyn x)))))
end;;
\end{code}

\oleg{Compare the output of some of the ctest with the corresponding
ptest. Use power?}

One should note that the partial evaluator has fully reused the code
of the compiler and the interpreter. It is truly the `composition' of the
interpreter and the compiler.  \jacques{Perhaps cite a paper of Thiemann
here, cogen in 6 lines?}

\jacques{Comment on the various nice generic things in the implementation,
which are a little silly in this small version, but would really pay off
in something bigger.  See the abstract interpretation hard at work!
See the monoid-driven simplifications!}
\jacques{Also need to note that the type of the above is not quite
the original Symantics (although the code has the right operational
semantics).  We need to make some adjustments, as per below.}

@@

\begin{code}
  ('c, int   ) repr_pe = VI of int
                       | E of ('c, int) code
  ('c, bool  ) repr_pe = VB of bool
                       | E of ('c, bool) code
  ('c, 'a->'b) repr_pe = VF of ('c,'a) repr -> ('c,'b) repr
                       | E of ('c, 'a->'b) code
\end{code}

@@

So we have to augment the full signature of Symantics (from that of
sec.~\ref{encoding}) to
\begin{code}
module type Symantics = sig
  type ('c,'sv,'dv) repr
  val int  : int  -> ('c,int,int) repr
  val bool : bool -> ('c,bool,bool) repr
  val add  : ('c,int,int) repr -> 
      ('c,int,int) repr -> ('c,int,int) repr
  val mul  : ('c,int,int) repr -> 
      ('c,int,int) repr -> ('c,int,int) repr
  val leq  : ('c,int,int) repr -> 
      ('c,int,int) repr -> ('c,bool,bool) repr
  val if_  : ('c,bool,bool) repr ->
             (unit -> ('c,'sa,'da) repr) ->
             (unit -> ('c,'sa,'da) repr) -> ('c,'sa,'da) repr 
  val lam : (('c,'sa,'da) repr -> ('c,'sb,'db) repr)
    -> ('c,(('c,'sa,'da) repr -> ('c,'sb,'db) repr),'da->'db) repr
  val app : ('c,(('c,'sa,'da) repr -> ('c,'sb,'db) repr),'da->'db) repr
    -> ('c,'sa,'da) repr -> ('c,'sb,'db) repr
  val fix : (('c,(('c,'sa,'da) repr -> ('c,'sb,'db) repr) as 's,'da->'db) repr 
             -> ('c,'s,'da->'db) repr)  -> ('c,'s,'da->'db) repr
end
\end{code}

\jacques{We talk a lot about GADT, but show none.  I think we need to
show a bit of Haskell, to explain what we are talking about.}

@@

$$ 
  (\ZZ \rightarrow r \ZZ) \rightarrow
  (\BB \rightarrow r \BB) \rightarrow
  \forall \alpha \beta. (r \alpha \rightarrow r \beta)
      \rightarrow r (\alpha\rightarrow\beta) \rightarrow ... r \tau
$$
% (Int -> r Int) ->
% (Bool -> r Bool) ->
% (forall alpha beta. (r alpha -> r beta) -> r (alpha->beta)) -> ... r tau

@@

Taha et al (cite different paper) define a self-interpreter si to be an
object-language term such that, for any object-language term e,

@@

Walid's PEPM07 (Concoqtion paper)? At leats at the Concoqtion
presentation at WG2.11, Emir used tagless interpreter as the main and
only example. \jacques{but at the PEPM07 presentation, Walid did not
quite do it that way.  He presented Concoqtion more as an exploration
of (an obvious but difficult) part of the design space for programming
languages.}

@@

Regarding our implementation of the type-level function |'sv -> 'dv|.
Gibbons et al paper on
the implementations of typecase may be relevant. I have not read it,
but I have the feeling some of their typecase designs may be
relevant (at least to cite). 
However, the typecase paper relies on having a 
bijection between types -- our |'sv| and |'dv| in this case, which is not 
our situation.  We have a one-sided view of things, where |'sv -> 'dv| is 
easy, and |'dv| "perfectly reflects" an |'sv|, but you still can't get at 
that |'sv|.  \jacques{Could we say that we have a meta-level Galois
injection?  We have one direction completely covered at the object
level (and so easy to lift), but 'dv to 'sv can only be done by
an outside mechanism}

@@

We show how to do many things without tags, in a simple type system, 
essentially by avoiding any (inductive) types.

\jacques{Should we mention that algebraic type tags give us a partly
intensional representation of terms.  This nicely allows pattern matching,
but as sits inside an extensional theory, ensuring that all traces 
of these tags disappear seems quite difficult without a very powerful
type system (enter GADTs).}
